---
title: "Coursera Data Science Courses Projects Reproducible Research Week 2 Project"
author: "Andualem Bekele"
date: "January 19, 2018"
output: html_document
---

##Loading and preprocessing the data

###   Loadind the data file
 

```{r}
setwd("C:/Users/abekele/Documents/Corsera/Reproducible Research")

```

####Looking at the five rows of the data  - summary 


```{r}


activitydata <- read.csv("activity.csv")
summary(activitydata)

```

```{r}

head(activitydata)

```

##What is mean total number of steps taken per day?

###For this part of the assignmentthe missing values in the dataset are ignored.

###1. Calculate the total number of steps taken per day

```{r}

steps <- aggregate(activitydata$steps, by = list(Date = activitydata$date), FUN = sum)
library(ggplot2)
names(steps)[names(steps) == "x"] <- "Total"
temp <- as.Date(steps$Date, "%Y-%m-%d")
steps$Date <- format(temp, format = "%m-%d")
head(steps)
```

###2.Make a histogram of the total number of steps taken each day.

```{r}
hist1 <- ggplot(data = na.omit(steps), aes(Total)) + 
    geom_histogram(binwidth = 1500, colour = "red") +
    xlab("Total Number of Steps Taken Each Day") +
    ylab("Count") +
    ggtitle("Total Number of Steps Taken Each Day")
print(hist1)

```

###3.Calculate and report the mean and median of the total number of steps taken per day

```{r}

mean(na.omit(steps$Total))
```
```{r}
median(na.omit(steps$Total))
```

###What is the average daily activity pattern?

###4.Make a time series plot of 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis) 

```{r}
five_min_steps <- aggregate(steps ~ interval, data = activitydata, FUN =mean)
TimeSeries1 <- ggplot(data = five_min_steps, aes(x = interval,  y = steps)) + 
    geom_line(col = "blue") +
    xlab("Time Intervals (5 Minutes is an unit)") + 
    ylab("Total Number of Steps") +
    ggtitle("Average Number of Steps Taken of the 5-Minute Interval")
print(TimeSeries1)
```



## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

Find the row with maximum number of steps and look at the 5 minute interval of that row

```{r}
head(five_min_steps)
```

```{r}

five_min_steps[which(five_min_steps$steps == max(five_min_steps$steps)),]
```

##Replacing missing values

Note that there are a number of days/intervals where there are missing values (coded as ????????). The presence of missing days may introduce bias into some calculations or summaries of the data.

####Calculate and report the total number of missing values in the dataset  (i.e. the total number of rows with NAs)

```{r}
sapply(X = activitydata, FUN = function(x) sum(is.na(x)))
```

####Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc. I will use the mean for that 5 -minute interval to replace all the missing values in the dataset. At the end, I will check if all the NAs have been replaced

```{r}
library(dplyr)
```

```{r}
replacewithmean <- function(num) replace(num, is.na(num), mean(num, na.rm = TRUE))
meanofday <- (activitydata %>% group_by(interval) %>% mutate(steps = replacewithmean(steps)))
head(meanofday)
```
```{r}
sum(is.na(meanofday))
```

###Create a new dataset with the missing data replaced with the mean.

```{r}
newdataset <- as.data.frame(meanofday)
head(newdataset)

```

```{r}
summary(newdataset)
```

####Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps? We make a histogram of the total number of steps taken each day first by using the new version dataset

```{r}
new_steps <- aggregate(newdataset$steps, by = list(newdataset$date), FUN = sum)
names(new_steps)[names(new_steps) == "x"] <- "Total"
names(new_steps)[names(new_steps) == "Group.1"] <- "Date"
hist2 <- ggplot(data = new_steps, aes(Total)) + 
  geom_histogram(binwidth = 1500, colour = "red")+
    xlab("Total Number of Steps Taken Each Day") +
    ylab("Count") +
    ggtitle("Total Number of Steps Taken Each Day with replaced values")
print(hist2)

```


compare the two plots

```{r}

print(hist1)
```

compare the two means
```{r}
mean(na.omit(steps$Total))
median(na.omit(steps$Total))
mean(new_steps$Total)
median(new_steps$Total)
```

##Are there differences in activity patterns between weekdays and weekends?

Dataset with the replaced missing values is used 

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.


```{r}
newdataset$WkendOrWkday <- ifelse(weekdays(as.Date(newdataset$date)) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "Weekday", "Weekend")
head(newdataset)

```
####Make a panel plot containing a time series plot (i.e. ???????????????? = "????") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.


```{r}

newdataset <- (newdataset %>% group_by(interval, WkendOrWkday) %>% summarise(Mean = mean(steps)))
ggplot(newdataset, mapping = aes(x = interval, y = Mean)) + geom_line(col = "blue") +
    facet_grid(WkendOrWkday ~.) + xlab("Interval") + ylab("Mean of Steps") +
    ggtitle("Average Number of Steps in Each Interval")
```

Weekend activity patterns differ from weekdays paterns. 